Calendar-based graphics for visualising people's daily schedules

### 2017-07-12 main.Rmd
Added polar graph
Added to boxplots

This work was originally motivated by studying foot traffic in the city of Melbourne [@ped]. There have been 43 installed sensors counting pedestrians every hour across the downtown area until the end of 2016 (see Figure \ref{fig:ped-map}). The dataset can shed lights into understanding people's daily schedules, or assisting administration and business planning. We start off with the conventional time series plot to take a glimpse at such data. Small multiples, shown as Figure \ref{fig:time-series-plot}, give an overall picture of foot traffic at a selection of 3 sensors. Figure \ref{fig:facet-time}, complemented to the Figure \ref{fig:time-series-plot}, provides another aspect of temporal patterns in the more detailed level by faceting day of the week. It lends itself to a number of issues that make exploratory data visualisation challenging in many temporal-context applications involving human behaviours:

@@

Figure \ref{fig:flinders-2016} shows the line glyphs organised in the monthly calendar over the year of 2016. This is achieved by the \code{frame_calendar} function computing the new coordinates according to the input data variables; in turn the rearranged data values are plotted using the \pkg{ggplot2} package [@R-ggplot2], which is the implementation of the grammar of graphics (@wilkinson2006grammar and @wickham2010layered).

@@

```
frame_calendar(
  data, x, y, date, calendar = "monthly", dir = "h", sunday = FALSE, 
  nrow = NULL, ncol = NULL, polar = FALSE, scale = "fixed"
)
```

@@

@@

Regarding the monthly calendar, the major reference lines separate every month panel and the minor ones separate every cell, represented by the think and thin lines respectively. The major reference lines are placed surrounding every month block: for each $m$, the vertical lines are determined by $\min{(x)}$ and $\max{(x)}$; for each $n$, the horizontal lines are given by $\min{(y)}$ and $\max{(y)}$. The minor reference lines are placed on the left side of every cell: for each $i$, the vertical division is $\min{(x)}$; for each $j$, the horizontal is $\min{(y)}$.

@@

```{r boxplot, fig.height = 6.5, fig.cap = "Side-by-side boxplot of hourly counts across all the 43 sensors in December of 2016, with the superimposing hourly averages. It shows the hourly distribution in the city as a whole. Christmas stands out as the quietest day in December."}
```

>>>>

This work was originally motivated by studying foot traffic in the city of Melbourne [@ped]. There have been 43 installed sensors counting pedestrians every hour across the downtown area until the end of 2016 (see Figure \ref{fig:ped-map}). The dataset can shed lights into understanding people's daily schedules, or assisting administration and business planning. We start off with the conventional time series plot to catch a glimpse of such data. Small multiples, shown as Figure \ref{fig:time-series-plot}, give an overall picture of foot traffic at a selection of 3 sensors. Figure \ref{fig:facet-time}, complemented to the Figure \ref{fig:time-series-plot}, provides another aspect of temporal patterns in the more detailed level by faceting day of the week. It lends itself to a number of issues that make exploratory data visualisation challenging in many temporal-context applications involving human behaviours:

@@

Figure \ref{fig:flinders-2016} shows the line glyphs framed in the monthly calendar over the year of 2016. This is achieved by the \code{frame_calendar} function computing the new coordinates according to the input data variables; in turn the rearranged data values are plotted using the \pkg{ggplot2} package [@R-ggplot2], which is the implementation of the grammar of graphics (@wilkinson2006grammar and @wickham2010layered).

@@

```
frame_calendar(
  data, x, y, date, calendar = "monthly", dir = "h", sunday = FALSE, 
  nrow = NULL, ncol = NULL, polar = FALSE, scale = "fixed",
  width = 0.95, height = 0.95
)
```

@@

```{r flinders-polar, fig.height = 6.5, fig.cap = "Line gplyphs on polar coordinates."}
```

@@

Regarding the monthly calendar, the major reference lines separate every month panel and the minor ones separate every cell, represented by the thick and thin lines respectively. The major reference lines are placed surrounding every month block: for each $m$, the vertical lines are determined by $\min{(x)}$ and $\max{(x)}$; for each $n$, the horizontal lines are given by $\min{(y)}$ and $\max{(y)}$. The minor reference lines are placed on the left side of every cell: for each $i$, the vertical division is $\min{(x)}$; for each $j$, the horizontal is $\min{(y)}$.

@@

The newly computed coordinates not only work with the simple geoms such as lines and points, but also boxplot and other complex geoms. Figure \ref{fig:boxplot} uses the side-by-side boxplots with the loess smooth line superimposed as an example. It shows the distribution of hourly counts across all the 43 sensors.

```{r boxplot, fig.height = 6.5, fig.cap = "Side-by-side boxplot of hourly counts across all the 43 sensors in December of 2016, with the superimposing loess smooth line for each day. It shows the hourly distribution in the city as a whole. Christmas stands out as the quietest day in December."}
```

In general, unimodality features the non-work days whereas bimodality on work days. The last week of December is just the holiday season: people are off work on the day before Christmas, go shopping on the Boxing day, and hanging out for the firework in the New Year Evening.

### 2017-07-12 main.tex
Added polar graph
Added to boxplots

This work was originally motivated by studying foot traffic in the city
of Melbourne \citep{ped}. There have been 43 installed sensors counting
pedestrians every hour across the downtown area until the end of 2016
(see Figure \ref{fig:ped-map}). The dataset can shed lights into
understanding people's daily schedules, or assisting administration and
business planning. We start off with the conventional time series plot
to take a glimpse at such data. Small multiples, shown as Figure
\ref{fig:time-series-plot}, give an overall picture of foot traffic at a
selection of 3 sensors. Figure \ref{fig:facet-time}, complemented to the
Figure \ref{fig:time-series-plot}, provides another aspect of temporal
patterns in the more detailed level by faceting day of the week. It
lends itself to a number of issues that make exploratory data
visualisation challenging in many temporal-context applications
involving human behaviours:

@@

Figure \ref{fig:flinders-2016} shows the line glyphs organised in the
monthly calendar over the year of 2016. This is achieved by the
\code{frame_calendar} function computing the new coordinates according
to the input data variables; in turn the rearranged data values are
plotted using the \pkg{ggplot2} package \citep{R-ggplot2}, which is the
implementation of the grammar of graphics (\citet{wilkinson2006grammar}
and \citet{wickham2010layered}).

@@

\begin{verbatim}
frame_calendar(
  data, x, y, date, calendar = "monthly", dir = "h", sunday = FALSE, 
  nrow = NULL, ncol = NULL, polar = FALSE, scale = "fixed"
)
\end{verbatim}

@@

@@

Regarding the monthly calendar, the major reference lines separate every
month panel and the minor ones separate every cell, represented by the
think and thin lines respectively. The major reference lines are placed
surrounding every month block: for each \(m\), the vertical lines are
determined by \(\min{(x)}\) and \(\max{(x)}\); for each \(n\), the
horizontal lines are given by \(\min{(y)}\) and \(\max{(y)}\). The minor
reference lines are placed on the left side of every cell: for each
\(i\), the vertical division is \(\min{(x)}\); for each \(j\), the
horizontal is \(\min{(y)}\).

@@

\caption[Lag scatterplot on the daily calendar]{Lag scatterplot on the daily calendar. Lagged hourly counts are plotted against current ones to demonstrate the autocorrelation at the order of 1. The positive association between the one-hour-ahead count and the present regardless of the data values takes place on non-working days.}\label{fig:scatterplot}
\end{figure}
\end{CodeChunk}

@@

\caption[Side-by-side boxplot of hourly counts across all the 43 sensors in December of 2016, with the superimposing hourly averages]{Side-by-side boxplot of hourly counts across all the 43 sensors in December of 2016, with the superimposing hourly averages. It shows the hourly distribution in the city as a whole. Christmas stands out as the quietest day in December.}\label{fig:boxplot}
\end{figure}
\end{CodeChunk}

>>>>

This work was originally motivated by studying foot traffic in the city
of Melbourne \citep{ped}. There have been 43 installed sensors counting
pedestrians every hour across the downtown area until the end of 2016
(see Figure \ref{fig:ped-map}). The dataset can shed lights into
understanding people's daily schedules, or assisting administration and
business planning. We start off with the conventional time series plot
to catch a glimpse of such data. Small multiples, shown as Figure
\ref{fig:time-series-plot}, give an overall picture of foot traffic at a
selection of 3 sensors. Figure \ref{fig:facet-time}, complemented to the
Figure \ref{fig:time-series-plot}, provides another aspect of temporal
patterns in the more detailed level by faceting day of the week. It
lends itself to a number of issues that make exploratory data
visualisation challenging in many temporal-context applications
involving human behaviours:

@@

Figure \ref{fig:flinders-2016} shows the line glyphs framed in the
monthly calendar over the year of 2016. This is achieved by the
\code{frame_calendar} function computing the new coordinates according
to the input data variables; in turn the rearranged data values are
plotted using the \pkg{ggplot2} package \citep{R-ggplot2}, which is the
implementation of the grammar of graphics (\citet{wilkinson2006grammar}
and \citet{wickham2010layered}).

@@

\begin{verbatim}
frame_calendar(
  data, x, y, date, calendar = "monthly", dir = "h", sunday = FALSE, 
  nrow = NULL, ncol = NULL, polar = FALSE, scale = "fixed",
  width = 0.95, height = 0.95
)
\end{verbatim}

@@

{\centering \includegraphics[width=\textwidth]{figure/flinders-polar-1}

\caption[Line gplyphs on polar coordinates]{Line gplyphs on polar coordinates.}\label{fig:flinders-polar}
\end{figure}
\end{CodeChunk}

@@

Regarding the monthly calendar, the major reference lines separate every
month panel and the minor ones separate every cell, represented by the
thick and thin lines respectively. The major reference lines are placed
surrounding every month block: for each \(m\), the vertical lines are
determined by \(\min{(x)}\) and \(\max{(x)}\); for each \(n\), the
horizontal lines are given by \(\min{(y)}\) and \(\max{(y)}\). The minor
reference lines are placed on the left side of every cell: for each
\(i\), the vertical division is \(\min{(x)}\); for each \(j\), the
horizontal is \(\min{(y)}\).

@@

\caption[Lag scatterplot in the daily calendar]{Lag scatterplot in the daily calendar. Lagged hourly counts are plotted against current ones to demonstrate the autocorrelation at the order of 1. The positive association between the one-hour-ahead count and the present regardless of the data values takes place on non-working days.}\label{fig:scatterplot}
\end{figure}
\end{CodeChunk}

The newly computed coordinates not only work with the simple geoms such
as lines and points, but also boxplot and other complex geoms. Figure
\ref{fig:boxplot} uses the side-by-side boxplots with the loess smooth
line superimposed as an example. It shows the distribution of hourly
counts across all the 43 sensors.

@@

\caption[Side-by-side boxplot of hourly counts across all the 43 sensors in December of 2016, with the superimposing loess smooth line for each day]{Side-by-side boxplot of hourly counts across all the 43 sensors in December of 2016, with the superimposing loess smooth line for each day. It shows the hourly distribution in the city as a whole. Christmas stands out as the quietest day in December.}\label{fig:boxplot}
\end{figure}
\end{CodeChunk}

In general, unimodality features the non-work days whereas bimodality on
work days. The last week of December is just the holiday season: people
are off work on the day before Christmas, go shopping on the Boxing day,
and hanging out for the firework in the New Year Evening.
